---
title: "Hierarchical Time Series with hts"
description: |
  This is just a quick reproduction of the items discussed in the 
  hts package. This allows for hierarchical time series which is 
  an important feature when looking at data that take a hierarchical
  format like counties within a state or precincts within counties within states.
categories: time series
date: 2018-10-28
---


## Mission

The mission is to reproduce the figures in the following article:

<https://cran.r-project.org/web/packages/hts/vignettes/hts.pdf>

## Load Required Libraries

```{r}
library(hts)
```

### Note on Formatting Data

This is important to understand in regard to how to format the data.

```{r}
# bts is a time series matrix containing the bottom-level series
# The first three series belong to one group, and the last two
# series belong to a different group
# nodes is a list containing the number of child nodes at each level.
bts <- ts(5 + matrix(sort(rnorm(500)), ncol=5, nrow=100))

y <- hts(bts, nodes=list(2, c(3, 2)))
```


## The Data

```{r}
infantgts
```

Use a bottom's up forcasting method. There are several options from which to choose:

- **comb** optimal combination
- **bu** bottom's up
- **mo** middle out forecast
- **tdgsa** Top-down forecasts based on the average historical proportions (Gross-Sohl method A)
- **tdgsf** same as above but with averages
- **tdfp** top down forecast using proportions

```{r}
# Forecast 10-step-ahead using the bottom-up method
infantforecast <- forecast(infantgts, h=10, method="bu")
```


Now we can plot them:

```{r}
# plot the forecasts including the last ten historical years
plot(infantforecast, include=2)

```



```{r}
allts_infant <- aggts(infantgts)
allf <- matrix(nrow=10, ncol=ncol(allts_infant))
# Users can select their preferred time-series forecasting method
# for each time series
for(i in 1:ncol(allts_infant)){
  allf[,i] <- forecast(auto.arima(allts_infant[,i]), h=10, PI=FALSE)$mean
}

allf <- ts(allf, start=2004)
# combine the forecasts with the group matrix to get a gts object
g <- matrix(c(rep(1, 8), rep(2, 8), rep(1:8, 2)), nrow = 2, byrow = T)
y.f <- combinef(allf, groups = g)
```

```{r}
# set up the training sample and testing sample
data <- window(infantgts, start=1933, end=1993)
test <- window(infantgts, start=1994, end=2003)
forecast <- forecast(data, h=10, method="bu")
# calculate ME, RMSE, MAE, MAPE, MPE and MASE
accuracy.gts(forecast, test)
```

```{r}
fcasts <- forecast(htseg1, h = 10, method = "comb", fmethod = "arima",
weights = "sd", keep.fitted = TRUE, parallel = TRUE)
```

