---
title: "Quick Review of Julia"
description: |
  Julia - runs like FORTRAN and Write Like Haskell or where I try to build a package in Julia an enjoyed it.
date: 2021-06-26
base_url: https://michaeldewittjr.com
twitter:
  creator: "@medewittjr"
categories: [Julia, Scientific Computing, Package Development]
link-citations: true
draft: true
---


```{r setup, include=FALSE}
library(JuliaCall)
#JuliaCall::julia_setup(JULIA_HOME = "/usr/local/bin/julia/")
```

## Why Julia?

Julia has been on my radar to learn for a while (since maybe 2017-18).
I love the idea of writing really fast code once without trying to translate operations into C++ with Rcpp. 

::: {.column-margin}
This is not to take away from Rcpp. It truly is amazing.
:::

Additionally, I have been doing a lot more package development at work to facilitate workflows and performant coding in general and packages with compiled code is always harder -- even if I build the binaries.

The other side of 



<aside>
```{r}
library(ggplot2)
ggplot(mtcars, aes(hp, mpg)) + geom_point() + geom_smooth()
```
</aside>

```julia

2+2;

```


